# The parallel R package

Although R was not built for parallel computing, multiple ways of parallelizing your R code exist. One of these is the `parallel` package. This R package, shipped with base R, provides various functions to parallelize R code using [embarrassingly parallel computing](https://en.wikipedia.org/w/index.php?title=Embarrassingly_parallel&oldid=1136401514){target="_blank"}, i.e., a divide-and-conquer-type strategy. The basic idea is to start multiple R sessions (usually called child processes), connect the main session with those, and send them instructions. This section goes over a common workflow to work with R's parallel.

## Parallel workflow

(Usually) We do the following:

1.  Create a `PSOCK/FORK` (or other) cluster using `makePSOCKCluster`/`makeForkCluster`
    (or `makeCluster`)
    
2.  Copy/prepare each R session (if you are using a `PSOCK` cluster):

    a.  Copy objects with `clusterExport`

    b.  Pass expressions with `clusterEvalQ`

    c.  Set a seed

3.  Do your call: `parApply`, `parLapply`, etc. 

4.  Stop the cluster with `clusterStop`


## Types of clusters: PSOCK

-   Can be created with `makePSOCKCluster`

-   Creates brand new R Sessions (so nothing is inherited from the master), e.g.
    
    ```r
    # This creates a cluster with 4 R sessions
    cl <- makePSOCKCluster(4)
    ```

-   Child sessions are connected to the master session via Socket connections

-   Can be created outside the current computer, **i.e.**, across multiple computers!

## Types of clusters: Fork

-   Fork Cluster `makeForkCluster`:

-   Uses OS [Forking](https://en.wikipedia.org/wiki/Fork_(system_call)),

-   Copies the current R session locally (so everything is inherited from
    the master up to that point).
    
-   Data is only duplicated if altered (need to double check when this happens!)

-   Not available on Windows.

Other `makeCluster`: passed to [**snow**](https://cran.r-project.org/package=snow)
(Simple Network of Workstations)


## Ex 1: Parallel RNG with `makePSOCKCluster`

::: {.callout-caution}
Using more threads than cores available on your computer is never a good idea. As a rule of thumb, clusters should be created using `parallel::detectCores() - 1` cores (so you leave one free for the rest of your computer.)
:::

```{r parallel-ex-psock, echo=TRUE}
# 1. CREATING A CLUSTER
library(parallel)
nnodes <- 4L
cl     <- makePSOCKcluster(nnodes)    
# 2. PREPARING THE CLUSTER
clusterSetRNGStream(cl, 123) # Equivalent to `set.seed(123)`
# 3. DO YOUR CALL
ans <- parSapply(cl, 1:nnodes, function(x) runif(1e3))
(ans0 <- var(ans))
```

Making sure it is reproducible

```{r parallel-ex-psock-cont, echo=TRUE}
# I want to get the same!
clusterSetRNGStream(cl, 123)
ans1 <- var(parSapply(cl, 1:nnodes, function(x) runif(1e3)))
# 4. STOP THE CLUSTER
stopCluster(cl)
all.equal(ans0, ans1) # All equal!
```

## Ex 2: Parallel RNG with `makeForkCluster`

In the case of `makeForkCluster`

```{r parallel-ex-fork, echo=TRUE, eval = TRUE}
# 1. CREATING A CLUSTER
library(parallel)
# The fork cluster will copy the -nsims- object
nsims  <- 1e3
nnodes <- 4L
cl     <- makeForkCluster(nnodes)    
# 2. PREPARING THE CLUSTER
clusterSetRNGStream(cl, 123)
# 3. DO YOUR CALL
ans <- do.call(cbind, parLapply(cl, 1:nnodes, function(x) {
  runif(nsims) # Look! we use the nsims object!
               # This would have fail in makePSOCKCluster
               # if we didn't copy -nsims- first.
  }))
(ans0 <- var(ans))
```

Again, we want to make sure this is reproducible

```{r parallel-ex-fork-cont, echo=TRUE}
# Same sequence with same seed
clusterSetRNGStream(cl, 123)
ans1 <- var(do.call(cbind, parLapply(cl, 1:nnodes, function(x) runif(nsims))))
ans0 - ans1 # A matrix of zeros
# 4. STOP THE CLUSTER
stopCluster(cl)
```

<text style="color:white;">Well, if you are a Mac-OS/Linux user, there's a more straightforward way of doing this...</text>


## Ex 3: Parallel RNG with `mclapply` (Forking on the fly)

In the case of `mclapply`, the forking (cluster creation) is done on the fly!

```{r parallel-ex-mclapply, echo=TRUE, eval = TRUE}
# 1. CREATING A CLUSTER
library(parallel)
# The fork cluster will copy the -nsims- object
nsims  <- 1e3
nnodes <- 4L
# cl     <- makeForkCluster(nnodes) # mclapply does it on the fly
# 2. PREPARING THE CLUSTER
set.seed(123) 
# 3. DO YOUR CALL
ans <- do.call(cbind, mclapply(1:nnodes, function(x) runif(nsims)))
(ans0 <- var(ans))
```

Once more, we want to make sure this is reproducible

```{r parallel-ex-mclapply-cont, echo=TRUE}
# Same sequence with same seed
set.seed(123) 
ans1 <- var(do.call(cbind, mclapply(1:nnodes, function(x) runif(nsims))))
ans0 - ans1 # A matrix of zeros
# 4. STOP THE CLUSTER
# stopCluster(cl) no need of doing this anymore
```