# Notes about scoping in C++

This chapter will present a more advanced but important topic in C++: RAII and scoping. RAII stands for Resource Acquisition Is Initialization, and it is a programming idiom that ensures that resources are properly released when they are no longer needed. From the perspective of an R user, this provides a nice opportunity for scoping in C++. Let's take a look with a simple example (which is also available in [scoping-cpp.hpp](./scoping-cpp.hpp)):

```{Rcpp}
#| label: scoping-cpp
#| cache: true
class Agent {
public:
    int fun();
};

// Declaration of the Model class
class Model {
private:
    Agent agent_;
    int x_;

public:
    inline static Model * instance_ = nullptr;
    virtual int run();
    Model(int x) : agent_(Agent()), x_(x) {};
    static Model & get();
    int get_x();
};

// Implementation of the Model methods
inline int Model::run() {
    Model::instance_ = this;
    return agent_.fun();
};

inline Model & Model::get() {return *instance_;};
inline int Model::get_x() {return x_;};

// Implementation of the Agent method
inline int Agent::fun() {return Model::get().get_x();};

// [[Rcpp::export]]
int test_scoping_cpp() {

    // Create an instance of Model
    Model Model(42);

    // Run the Model
    return Model.run();
}

// [[Rcpp::export]]
bool is_it_on() {
    return Model::instance_ != nullptr;
}
```

Checking whether it works or not:

```{r}
#| label: checking-test-scoping
test_scoping_cpp()
```

It does! Let's split this into pieces. First of all, the `Agent` class does not have explicit access to the `Model` class; in no place of its definition has it as a member. The key lies in the `static` function `Model::get()`, which can be call by any function within the call stack without having to access to `Model`.

```{r}
is_it_on()
```

Now, this is not entirely safe, as the variable `instance_` would still be available even after calling the `Model::run()` function. To address this, we can use an auxiliary class that resets the `instance_` to `nullptr`:

```{Rcpp}
#| label: scoping-cpp-scope
#| cache: true
#include <Rcpp.h>
class Agent {
public:
    int fun();
};

class Scope;

// Declaration of the Model class
class Model_s {
    friend Scope;
private:
    Agent agent_;
    int x_;

public:
    inline static Model_s * instance_ = nullptr;
    virtual int run();
    Model_s(int x) : agent_(Agent()), x_(x) {};
    static Model_s & get();
    int get_x();
};

// Declaration & Implementation of the Scope method
class Scope {
private:
    Model_s * prev_;
public:
    explicit Scope(Model_s * model) : prev_(Model_s::instance_) {
        Model_s::instance_ = model;
    };
    ~Scope() {
        Model_s::instance_ = prev_;
    };
};

// Implementation of the Model_s methods
inline int Model_s::run() {
    Scope scope(this);
    return agent_.fun();
};

inline Model_s & Model_s::get() {return *instance_;};
inline int Model_s::get_x() {return x_;};

// Implementation of the Agent method
inline int Agent::fun() {return Model_s::get().get_x();};

// [[Rcpp::export]]
int test_scoping_cpp2() {

    // Create an instance of Model
    Model_s Model(42);

    // Run the Model
    return Model.run();
}

// [[Rcpp::export]]
bool is_it_on2() {
    return Model_s::instance_ != nullptr;
}
```

The simple class `Scope` provides a nice way of controlling the duration of this static variable:

```cpp
class Scope {
private:
    Model_s * prev_;
public:
    explicit Scope(Model_s * model) : prev_(Model_s::instance_) {
        Model_s::instance_ = model;
    };
    ~Scope() {
        Model_s::instance_ = prev_;
    };
};
```

Once it is invoked, it sets the value for the static member of `Model_s` for the duration of `Scope`. Once `Model_s::run()` exists, the destructor of `Scope` resets the value of `Model_s::instance_` to its previous value, `nullptr`, in this case. Nonetheless, this provides a way to allow nested calls setting the proper value of `instance_` as the callstack grows. Here is the result of the new implementation

```{r}
test_scoping_cpp2() # Checking it works
is_it_on2()         # And that it is turned off
```

::: {.callout-tip}
While implementing this, I was using `Model` for both code chunks in C++. Although it still allows to me compile, the symbol `Model` is shared by both implementations, which broke my code (the `is_it_on2()` function was returning `TRUE` when it was supposed to return `FALSE`).
:::