# Simulating pi (once more)

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, results = 'asis', warning = FALSE, comment = "")
cat2 <- function(..., lang = 'r') {
  cat("```", lang, "\n", sep = "")
  cat(...)
  cat("```\n")
}
```



This is the same old example that lots of people (including me) have been using
to illustrate parallel computing with R. The example is very simple, we want to
approximate pi by doing some Monte Carlo simulations.

We know that the area of a circle is $A = \pi r^2$, which is equivalent to say
$\pi = A/r^2$, so, if we can approximate the Area of a circle, then we can
approximate $\pi$. How do we do this?

Using Monte Carlo experiments, we can approximate the probability of a random point 
$x$ falling within the unit circle using the following formula:

$$
\hat p = \frac{1}{n}\sum_i \mathbf{1}(x \in \mbox{Circle})
$$

This approximation, $\hat p$, multiplied by the area of the square containing the circle, which
has an area equal to $(2\times r)^2$, thus, we can finally write

$$
\hat \pi = \hat p \times (2\times r)^2 / r^2 = 4 \hat p
$$

```{r}
#| label: simplot
#| echo: false
#| out-width: '50%'
#| fig-width: 5
#| fig-height: 5.5
#| fig-align: center
#| fig-cap: 10,000 random points drawn within the unit circle.
set.seed(2312)
x <- matrix(runif(1e4*2, min = -1, max = 1), ncol=2)
plot(
  x, type = "p",
  col  = adjustcolor(
    c("steelblue", "tomato")[(sqrt(rowSums(x^2)) < 1) + 1],
    alpha.f = .5
    ),
  xlab = expression(symbol(x)),
  ylab = expression(symbol(y)),
  pch  = 19
  )
```


## Submitting jobs to Slurm

The main way that we will be working is by submitting jobs using the `sbatch` function.
This function takes as a main argument a bash file with the program to execute.
In the case of R, a regular bash file looks something like this:

```{r}
cat2(readLines("01-sapply.slurm"), sep="\n", lang = 'bash')
```

This file has three components:

- **The Slurm flags `#SBATCH`**: These comment-like entries are used to pass Slurm options to the job. In this example, we are only specifying the options `job-name` and `time`. Other common options to include would be `account` and `partition`.

- **Loading R `module load R/4.2.2`**: Depending on your system's configuration, you may or may not need to load modules or run bash scripts before being able to run R. In this example, we are loading R version 4.2.2 using `LMod` (see previous section).

- **Executing the R script**: Finally, after specifying Slurm options and loading whatever needs to be loaded before executing R, we are using `RScript` to execute the program we wrote.

Submission is then done as follows:

```bash
sbatch 01-sapply.slurm
```

The following examples have two files, a bash script and an R script to be called
by Slurm.


### Case 1: Single job, single core job

The most basic way is submitting a job using the `sbatch` command. In this
case, you need to have 2 files: (1) An R script, and (2) a bash script. e.g.

The contents of the R script ([01-sapply.R](01-sapply.R){target="_blank"}) are:


```{r}
cat2(readLines("01-sapply.R"), sep="\n", lang = 'r')
```


The contents of the bashfile ([01-sapply.slurm](01-sapply.slurm){target="_blank"}) are:

```{r}
cat2(readLines("01-sapply.slurm"), sep="\n")
```
    
### Case 2: Single job, multicore job

Now, imagine that we would like to use more than one processor for this job,
using something like the `parallel::mclapply` function from the parallel package.[^mclapply]
Then, besides adapting the code, we need to tell Slurm that we are using
more than one core per task, as in the following example:

[^mclapply]: This function is sort of a wrapper of `makeForkcluster`. [Forking](https://en.wikipedia.org/w/index.php?title=Fork_(system_call)&oldid=1145232797){target="_blank"} provides a way to duplicate a process in the OS without replicating the memory, which is both faster and efficient. 

R script ([02-mclapply.R](02-mclapply.R){target="_blank"}):

```{r}
cat2(readLines("02-mclapply.R"), sep="\n")
```

Bashfile ([02-mclapply.slurm](02-mclapply.slurm){target="_blank"}):

```{r}
cat2(readLines("02-mclapply.slurm"), sep="\n", lang = "bash")
```

## Jobs with the slurmR package 

The `slurmR` R package [@slurmRpaper; @slurmRpkg] is a lightweight wrapper of Slurm. The package's main functions are the `*apply` family--mostly through Slurm job arrays--and the `makeSlurmCluster()`--which is a wrapper of `makePSOCKcluster`. 

This section will illustrate how to submit jobs using the `makeSlurmCluster()` function and `Slurm_sapply`. Furthermore, the last example demonstrates how we can skip writing Slurm scripts entirely using the `sourceSlurm()` function included in the package.

### Case 3: Single job, multinode job

In this case, there is no simple way to submit a multinodal job to Slurm... unless you use the [**slurmR**](https://github.com/USCbiostats/slurmR){target="_blank"} package.[^installingslurmr] In this example, we will combine `slurmR` with the parallel package's `parSapply` function to submit a multinodal job using the function `makeSlurmCluster()`. With it, `slurmR` will submit a job requesting `njobs` tasks (processors) that could span multiple nodes,[^notexactlymultinode] and create a Socket cluster out of it (like using `makePSOCKcluster`.) One thing to keep in mind is that Socket clusters are limited in the number of connections a single R session can span. You can read more about it [here](https://stat.ethz.ch/pipermail/r-sig-hpc/2012-May/001373.html){target="_blank"} and [here](https://github.com/HenrikBengtsson/Wishlist-for-R/issues/28){target="_blank"}.

[^installingslurmr]: See installation instructions [**here**](https://github.com/USCbiostats/slurmR#installation){target="_blank"}

[^notexactlymultinode]: Although possible, most multinode jobs will be allocated only if there are not enough threads within a single node. Remember Slurm does not run the jobs, but rather reserves computational resources for you to run it.




R script ([03-parsapply-slurmr.R](03-parsapply-slurmr.R){target="_blank"}):

```{r}
cat2(readLines("03-parsapply-slurmr.R"), sep="\n")
```

Bashfile ([03-parsapply-slurmr.slurm](03-parsapply-slurmr.slurm){target="_blank"}):

```{r}
cat2(readLines("03-parsapply-slurmr.slurm"), sep="\n", lang = "bash")
```


### Case 4: Multi job, single/multi-core

Another way to submit jobs is using [**job arrays**](https://slurm.schedmd.com/job_array.html). A job array is essentially a job that is repeated `njobs` times with the same configuration. The main difference between replicates is what you do with the `SLURM_ARRAY_TASK_ID` environment variable. This variable is defined within each replicate and can be used to make the "subjob" depending on that.

Here is a quick example using R

```r
ID <- Sys.getenv("SLURM_ARRAY_TASK_ID")
if (ID == 1) {
  ...[do this]...
} else if (ID == 2) {
  ...[do that]...
}
```

The `slurmR` R package makes submitting job arrays easy. Again, with the simulation
of pi, we can do it in the following way:

R script ([04-slurm_sapply.R](04-slurm_sapply.R){target="_blank"}):

```{r}
cat2(readLines("04-slurm_sapply.R"), sep="\n")
```

Bashfile ([04-slurm_sapply.slurm](04-slurm_sapply.slurm){target="_blank"}):

```{r}
cat2(readLines("04-slurm_sapply.slurm"), sep="\n", lang = "bash")
```


One of the main benefits of using this approach instead of the `makeSlurmCluster`
function (and thus, working with a SOCK cluster) are:

-  The number of jobs is not limited here (only by the admin, but not by R). 

-  If a job fails, then we can re-run it using `sbatch` once again (see example
   [here](https://github.com/USCbiostats/slurmR#example-2-job-resubmission){target="_blank"}).
   
-  You can check the individual logs of each process using the function `Slurm_lob()`.

-  You can submit the job and quit the R session without waiting for it to finalize.
   You can always read back the job using the function `read_slurm_job([path-to-the-temp])`
   

### Case 5: Skipping the .slurm file

The `slurmR` package has a function named `sourceSlurm` that can be used to avoid
creating the `.slurm` file. The user can add the SBATCH options to the top of 
the R script (including the `#!/bin/sh` line) and submit the job from within R
as follows:

R script ([05-sapply.R](05-sapply.R){target="_blank"}):

```{r}
cat2(readLines("05-sapply.R"), sep="\n")
```

From the R console (is OK if you are in the Head node)

```r
slurmR::sourceSlurm("05-sapply.R")
```

And voilá! A temporary bash file will be generated and used to submit the R script
to the queue.
